Compositional novelty of plant, fungal and bacterial communities across urban habitats

### 2025-03-17 doc/manuscript.Rmd
minor figure edits
submission UFUG

---
title: "Soil drivers of plant, fungal and bacterial community novelty in urban habitats of the city of Xixón, Spain"
output: 
  bookdown::github_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: journal-of-vegetation-science.csl
link-citations: yes
always_allow_html: true
---

@@

Running title: Urban drivers of plant, fungal and bacterial novelty

Eduardo Fernández-Pascual^1,a^ (<https://orcid.org/0000-0002-4743-9577>), Zuzana Ferencova^2^ (<https://orcid.org/0000-0002-0915-0916>),
Víctor González-García^1^ (<https://orcid.org/0000-0002-8949-7943>), Borja Jiménez-Alfaro^1^ (<https://orcid.org/0000-0001-6601-9597>)

^1^ Biodiversity Research Institute (IMIB), University of Oviedo - CSIC - Principality of Asturias, Mieres, Spain; ^2^ Jardín Botánico Atlántico, Gijón/Xixón, Spain

Correspondence: Eduardo Fernández-Pascual, Instituto Mixto de Investigación en Biodiversidad, Campus de Mieres, Edificio de Investigación, 5ª planta, c/ Gonzalo Gutiérrez Quirós s/n, E-33600 Mieres, Spain. Email: [fernandezpeduardo\@uniovi.es](mailto:fernandezpeduardo@uniovi.es){.email}. Telephone: +34985104781.

# Author contributions

Eduardo Fernández-Pascual: Conceptualization; Methodology; Investigation; Data Curation; Formal Analysis; Visualization; Writing – Original Draft Preparation; Writing – Review & Editing. Zuzana Ferencova: Conceptualization; Methodology; Investigation; Data Curation; Writing – Review & Editing. Víctor González-García: Investigation; Borja Jiménez-Alfaro: Methodology; Writing – Review & Editing.

# Acknowledgements

María García Fernández, Sara Gutiérrez Cruz and Abelardo Sigüenza Solís helped with the field surveys.

# Funding

Fundación Biodiversidad, Ministerio para la Transición Ecológica y el Reto Demográfico, Gobierno de España [XIXÓN ECORRESILIENTE, PRTR, NextGenerationEU]. Ayuntamiento de Gijón/Xixón, Jardín Botánico Atlántico [SV-23-GIJON-JBA].

# Conflict of interest statement

None declared.

# Data availability statement

Original datasets, R code for analysis and Rmarkdown code for creation of the manuscript can be accessed at Zenodo: https://zenodo.org/doi/10.5281/zenodo.11613571.

@@

In post-industrial landscapes affected by the phenomenon of urban contraction, land located in the heart of cities remains unused for long periods (Haase et al., 2014) . When the use of any urban plot ceases, it is colonized by plants from the regional flora capable of living in urban environments (Johnson et al., 2017) . These urban plants are adapted to disturbances and can grow in degraded soils; some even accumulate heavy metals, making them nature-based solutions for bioremediation of degraded lands (Song et al., 2019) . Species traits, along with the legacy of human impact, determine which regional flora species colonize which plots (Swan et al., 2021) . When the human legacy is intense, the plant communities that develop will be novel communities, different from the historical reference communities that existed before human intervention (Gedan and Fernández-Pascual, 2019) . The novelty of the communities can be reflected in their floristic, compositional and phylogenetic composition, including their proportion of exotic species (Lososová et al., 2012) . In any case, proper management of urban communities can turn them into resources with high ecological and social potential (Anderson and Minor, 2017). (Kowarik, 2018) : in the Ibero-Atlantic ecoregion, anthropogenic habitats can host up to a third of the ecoregional flora.

The ability of plants to colonize any habitat, including urban ones, depends largely on their seed regeneration niche, that is, on their dispersal, persistence, germination and establishment functions (Grubb, 1977, Saatkamp et al., 2019) . However, although it is known that urban species often share a series of functional and ecological traits (Aronson et al., 2007, Lososová et al., 2006) , little is known about their seed regeneration niche. Especially, information on its germination traits is lacking. In species living in non-anthropogenic environments, germination phenology is regulated by environmental factors (water, temperature, light) to ensure that germination occurs at the right time to maximize subsequent seedling survival (Carta et al., 2022a) . Compared to the large amount of data on the germination of species from non-anthropogenic habitats, little is known about what factors regulate the germination of urban species. Considering other habitats affected by humans (Fernández-Pascual et al., 2022) , it could be expected that urban seeds have their germination unrelated to macroclimatic signals and show tolerance to stress due to heat, drought or heavy metals. However, this hypothesis needs to be tested through a synthesis of the seed regeneration niche in urban environments.

Cities 
1. highly fragmented landscapes
2. mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use)
3. subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems
4. comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure
5. contains a multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix

Cities are comprised of a patchwork of paved and unpaved spaces, built and vacant land, and newly developed and obsolescent and/or abandoned buildings and infrastructure. They are mosaics of heterogeneous ecosystems (in terms of size, fragmentation, population density and land use) which are subject to different degrees of human interference, leading to a stepwise transformation of pristine ecosystems to novel urban ecosystems (Kowarik, 2005).

The “Four Nature approach” narrows down the variety of transformational stages in urban settings to four major types that can supply wilderness in urban region (Kowarik 1992, 2005). Each of the four types relates to nature in general but results from different trajectories in human-nature interactions:

Spontaneous vegetation appears in different habitat types in the urban matrix,. Multitude of vacant lots, railway sidings, utility easements, corridors between buildings and canal sides that are often overgrown with spontaneous vegetation, which are not coherently managed, and which seem to occupy an uncertain, interstitial niche in the urban matrix. The spontaneous herbs could provide the very interesting pool of species for ecological restoration of the urban area. Their ecological value has been clear since many years because their contribution to the ecosystem services in urban areas. Although their aesthetic value has been highlighted repeatedly, their acceptance by the residents in urban areas has been questioned. Many residents associate the presence of spontaneous plants with the neglect in maintenance by public (municipal) management service. Spontaneous native urban vegetation has commonly been described as demonstrating resilience (Ignatieva et al., 2000) and exhibiting adaptations to human disturbance (Lundholm and Marlin, 2006; Sukopp, 2004).

Urban species pool is comprised of native species present before the urban development, native species developed because the new urban ecosystem and introduced species by human activity. However, just some of them can adapt to urban environment.

IGS (informal green space) consists of any urban space with a history of strong anthropogenic disturbance that is covered at least partly with non-remnant, spontaneous vegetation (Del Tredici, 2010).

@@

Gijón/Xixón is a medium-sized coastal city (population c. 270,000 inhabitants) located in the Bay of Biscay, within the central Asturias metropolitan area of northern Spain (population c. 800,000 inhabitants). The city belongs to the Cantabrian Mixed Forests ecoregion [@RN4667], at the border between the temperate and Mediterranean climatic zones of Europe, and has a temperate oceanic climate with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1050 mm). Most of the city is built on alluvial and estuarine deposits, sand, or limestone bedrock. The potential pre-human vegetation would be dominated by deciduous mixed forests, together with riparian forests, wetlands and coastal dunes. The area has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. The city itself has been inhabited at least since Roman times [@RN5532]. During the medieval and early modern era the urban center was limited to the original walled city, with fishing as the main economic activity. Expansion beyond the original core started in the 18th and 19th centuries. Exposure to global trade networks and the mild climate turned the region into a hotspot for biological invasions, especially from the Americas and Asia [@RN5493]. In 1852, the city was connected by railway and became a major port for the regional mining and industrial centers. During the 20th century the city experienced significant industrial development (e.g. canning, coal, steel, shipbuilding) and rapid population growth, becoming the most populated city in the Principality of Asturias [@RN5534; @RN5533]. Industrial and urban growth brought environmental problems including loss of natural wetlands and air and watercourse pollution [@RN5535]. Like the rest of its metropolitan area, the city has been affected by recent post-industrial land-use changes which have left large areas with abandoned industrial sites (i.e. brownfields) [@RN5448; @RN5449].

@@

To survey urban biodiversity across habitats, we defined six habitats types (**Figure 1A**) representing different states within the four-nature classification of urban wilderness [@RN5445]: (1) forest patches < 10 km from the city center, i.e. type-1 nature in remnants of the pre-urban vegetation; (2) hay meadows, i.e. type-2 nature in remnants of human-made ecosystems resulting from agricultural habitat transformation; (3) parks, i.e. type-3 nature in managed green spaces created after habitat destruction; (4) roadsides, i.e. type-4 nature in relatively unmanaged strips of land alongside roads, railways, etc; (5) residential lots, i.e. type-4 nature in unmanaged vacant lots of previous residential use; and (6) industrial lots, i.e. type-4 nature in unmanaged vacant lots of previous  industrial use.

## Selection of sampling plots

During June 2023, we surveyed the city to establish five 4 x 4 m sampling plots per each of the six habitat types (n = 30 plots). The plots were distributed across the city (**Figure 1B**) and represented the variation of vegetation types found within each habitat type. In two plots per habitat (**Figure 1B**), we buried a datalogger (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil. Dataloggers recorded temperature once every hour and stayed on site for approximately one year, after which we retrieved them and downloaded their records to describe the thermal environment of the city and the habitats.

@@

From each of the 30 plots, we collected 500 g of soil for analysis of chemical composition and physical soil properties. We took the samples from 5 points in each plot (center + 4 cardinal points). We removed stones and other debris particles, sieved the soil in a 5 mm mesh and homogenized the sample in a plastic bag. We cleaned and sterilized all tools with water and ethanol between samples. We sent the samples for analysis at INEA - Escuela Universitaria de Ingeniería Agrícola, Universidad Pontificia de Comillas. Soil analysis included pH, texture, bulk density, electrical conductivity, organic matter content, total nitrogen, C/N ratio, active limestone, total carbonates, assimilable phosphorus, K, Ca, Mg, Na and heavy metals (Zn, Cr, Cu, Cd, Hg, Pb, Ni).

> Añadir detalles de los métodos de suelo (Zuzana)

@@

In each plot, we recorded all the spontaneous vascular plant species, as well as the cover (%) of each species in the plot. For plant species names we followed Euro+Med [@RN5044]. Additionally, from each of the 30 plots, we collected soil samples for DNA metabarcoding analysis of soil fungal and bacterial communities. We performed the sampling in sterile conditions, i.e. before entering the plot for any of the other tasks. In each plot, we took samples from the four cardinal points, 0.5 m inwards from the edge, at a depth of 5 cm. Before taking the samples, we removed the vegetation with a sterile tool. Then, we collected the soil samples in a 50 ml Falcon tube, mixed them, homogenized them with a lancet and finally stored the sample into a 2 ml Eppendorf tube which we immediately froze. We sent the frozen tube to AllGenetics & Biology SL (www.allgenetics.eu) for DNA metabarcoding analysis.

## DNA metabarcoding analysis

DNA analysis by AllGenetics included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi and 16S for bacteria); sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size < 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, taxonomic assignment, number of reads per taxon and generation of rarefaction curves.

*DNA isolation and quantification*. DNA was isolated from each soil sample using the DNeasy PowerSoil Pro DNA isolation kit (Qiagen) following the manufacturer’s instructions, and DNA was eluted in a final volume of 50 μL. In one sample the above-mentioned protocol did not yield a DNA extract of enough concentration, so DNA was isolated using the FastDNA Spin Kit for Soil (MPBio) following the manufacturer’s instructions, and DNA was eluted in a final volume of 70 μL. An extraction blank (Bex) was included in each round of DNA extraction, and treated as the plot samples to check for contamination. DNA concentration in each extract was quantified using the Qubit High Sensitivity dsDNA Assay (Thermo Fisher Scientific).

*DNA metabarcoding library preparation and sequencing*. For fungal library preparation, a fragment of the nuclear ITS1 region of around 280-300 bp (including the primer sequences) was amplified using the following primers: forward - ITS1f (5’ CTTGGTCATTTAGAGGAAGTAA 3’) [@RN5548]; reverse - ITS2 (5’ GCTGCGTTCTTCATCGATGC 3’) [@RN5549]. For prokaryotic library preparation, a fragment of the 16S rRNA gene of around 300 bp (including primer sequences) was amplified using the following primers: forward - 515F-Y (5’ GTGYCAGCMGCCGCGGTAA 3’) [@RN5536]; reverse - 806RB (5’ GGACTACNVGGGTWTCTAAT 3’) [@RN5537]. These primers also included the Illumina sequencing primer sequences attached to their 5’ ends. In the first amplification step, PCRs were carried out in a final volume of 12.5 μL, containing 1.25 μL of template DNA (1:2 diluted in PCR round 3; not diluted in PCR round 11), 0.5 μM of the primers, 3.13 μL of Supreme NZYTaq 2x Green Master Mix (NZYTech), and ultrapure water up to 12.5 μL. The fungal reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 35 cycles of 95 ºC for 30 s, 49 ºC for 45 s, 72 ºC for 45 s, and a final extension step at 72 ºC for 7 min. The bacterial reaction mixture was incubated as follows: an initial denaturation at 95 ºC for 5 min, followed by 25 cycles of 95 ºC for 30 s, 47 ºC for 45 s, 72ºC for 45 s, and a final extension step at 72 ºC for 7 min. The oligonucleotide indices that are required for multiplexing different libraries in the same sequencing pool were attached in a second amplification step with identical conditions but only 5 cycles and 60 ºC as the annealing temperature. A negative control that contained no DNA (BPCR) was included in every PCR round to check for contamination during library preparation. Library size was verified by running the libraries on 2 % agarose gels stained with GreenSafe (NZYTech) and imaging them under UV light. Then, libraries were purified using the Mag-Bind RXNPure Plus magnetic beads (Omega Bio-tek), following the instructions provided by the manufacturer. Finished libraries were pooled in equimolar amounts according to the results of a Qubit dsDNA HS Assay (Thermo Fisher Scientific) quantification. The pool was sequenced in a fraction of a NovaSeq PE250 flow cell (Illumina) aiming for a total output of 2 gigabases. Any potential traces of adapter dimers were removed using Cutadapt v3.5 [@RN5538]. Quality of the sequencing data was assesed with the software FastQC [@RN5539].

*Processing of sequencing data and inference of ASVs*. Cutadapt v3.5 was used to trim off non-biological DNA sequences (primers, indices, and sequencing adapters). Then, amplicon reads were processed using the DADA2 pipeline [@RN5541] implemented in QIIME 2 (release 2023.7) [@RN5540]. The first step in the DADA2 pipeline was to remove PCR primers and filter reads according to their quality. After checking the read quality profiles, forward reads were truncated at position 175, and reverse reads at position 131. The next step was to denoise and infer Amplicon Sequence Variants (ASVs). Error rates were learned from the dataset to denoise, using the parametric error model implemented in DADA2. Dereplication of the dataset (i.e. combination of all identical reads into unique reads) was carried out to reduce computational effort. These dereplicated forward and reverse reads were used to infer ASVs with the core sample inference algorithm [@RN5541]. Subsequently, corresponding R1 and R2 reads were merged into pairs with a minimum overlap of 12 identical base pairs. The final step in the DADA2 pipeline was to reduce the impact of artefacts (chimaeras) in the dataset. The final output of the DADA2 pipeline was a table containing the count of reads of every observed ASV in each sample, as well as a table listing all ASVs and their corresponding representative sequences.

*Taxonomic assignment*. Taxonomic assignment of each ASV was performed using a pre-trained classifier of the UNITE reference database for fungi [@RN5550] (release May 2021); and of the SILVA reference database for bacteria [@RN5542] (release 138.1 August 2020). To compare the representative sequences of the ASVs to the reference database, and compute the taxonomic assignment,the algorithm sklearn [@RN5543] implemented in QIIME 2 was applied, setting a minimum confidence level of 0.8. Several filters were applied to exclude: (1) singletons (i.e. ASVs containing only one member sequence in the whole data set); (2) ASVs occurring at a frequency below 0.01 % in each sample, to reduce the mistagging bias [@RN5544]; (3) from the bacterial dataset, non-prokaryotic ASVs such as eukaryotic sequences of plastid [@RN5545] and mitochondrial [@RN5546] origin; (4) those assigned only at domain (Bacteria) or kingdom (Fungi) level; (5) ASVs present in the negative controls that were more abundant in the negative controls than in the samples, indicating contamination [@RN5547].

*Alpha rarefaction plots*. Alpha rarefaction plots showed that the sequencing depth to reach saturation was 30,000 reads for the fungi and 22,000 reads for the bacteria, above the sequencing depth of all plots for fungi (mean = 114,919 reads; min = 79,045; max = 157,110) and bacteria (mean = 113,341 reads; min = 84,094; max = 171,404).

> Rarefaction

## Statistical analysis

@@

## Soil temperatures and properties across habitats

## Plant, fungal and bacterial diversity across habitats

> PERMANOVAs

> Analysis with species, genera and ASVs

> Analysis with presence absence

## Community novelty across habitats and taxonomic groups

## Soil drivers of novelty

>>>>

---
title: "Compositional novelty of plant, fungal and bacterial communities across urban habitats"
output: 
  bookdown::word_document2:
    reference_docx: template.docx
    number_sections: false
bibliography: refs.bib
csl: urban-forestry-and-urban-greening.csl
link-citations: yes
always_allow_html: true
---

@@

Running title: Compositional novelty across urban habitats

# Abstract (250 words)

We tested the hypothesis that cities are composed of different degrees of ecological novelty by studying four urban habitats with differing degrees of management and human legacy: park lawns, roadsides, residential vacant lots, and industrial vacant lots. We focused on community compositional novelty, by comparing the plant, fungal and bacterial species composition between urban habitats and two reference pre-urban habitats: forest remnants and hay meadows. We used a compositional novelty index based on multidimensional ordination, which is straightforward to calculate and only requires  species co-occurrence data for urban and reference habitats. As expected, (1) plant communities displayed the highest novelty and soil bacteria the lowest; (2) urban communities were markedly different from forest communities and relatively more similar to meadow communities; and (3) the degree of compositional novelty was highest in industrial vacant lots. On the contrary, managed park lawns, which we had expected to be highly novel, were relatively close to hay meadows. The lowest novelty among urban habitats was recorded in residential vacant lots, which had biological communities that more closely resembled those of pre-urban habitats. Our results highlight the effect of habitat type as a major driver of urban community composition and compositional novelty. These results suggest that city biodiversity can be enhanced by an integrative approach to the urban landscape that favors habitat heterogeneity by passive rewilding of managed park lawns, non-intervention on residential vacant lots, direct restoration of industrial vacant lots, and conservation of natural and agricultural habitat remnants as sources of native species.

# Keywords (7)

Urban biodiversity, vegetation, soil fungi, soil bacteria, DNA metabarcoding, soil pollution, novel ecosystems

# Highlights (3 to 5, 85 words each)

* A compositional novelty index based on multidimensional ordination is straightforward to calculate and only requires species co-occurrence data for urban and reference pre-urban habitats. 
* Novelty was the highest for plant communities and the lowest for soil bacteria. 
* Urban communities were markedly different from forest communities and relatively more like meadow communities.
* Compositional novelty was highest in industrial vacant lots, and lowest in residential vacant lots. 
* Park lawns were relatively close in composition to hay meadows.

@@

Urban development affects landscape composition and related biodiversity patterns worldwide [@RN5627]. Urbanization converts natural and agricultural habitats into a fine-scale mosaic of diverse urban habitats varying in size, fragmentation, past and present land uses, and degrees of human intervention [@RN5527]. At the same time, novel urban habitats experience climatic differences from the surrounding landscape, as well as high within-city microclimatic variation [@RN5601; @RN5603]. Urban soils likewise vary in permeability, compaction, organic matter content and human-concentrated chemicals [@RN5623; @RN5603].

Urban habitats are colonized by organisms from the regional species pool that are capable of living in urban environments [@RN2300]. The process of colonization requires species to pass through a series of hierarchical filters (biogeographical, abiotic, biotic and human) resulting in the formation of an urban species pool and the assembly of specialized urban communities [@RN5505; @RN2306]. Urbanization promotes specialized ruderal species and traits [@RN5613; @RN5622; @RN5679], and thus urban communities are rich in species that can tolerate urban disturbance, drought and pollution [@RN5603]. For this reason, the urban biota can be a source of nature-based solutions, such as the bioremediation of degraded soils [@RN5447; @RN5449] and ecosystem design [@RN5628; @RN5629]. Urban communities also host a high proportion of the regional biodiversity [@RN5600; @RN5682] and make up a major part of the supply of natural elements in densely populated landscapes [@RN5445]. Proper management of urban communities can turn them into assets with high ecological and social returns [@RN5446; @RN5676; @RN5678].

The diversity of urban communities can be characterized along two axes: naturalness and novelty [@RN5445]. Naturalness refers to the degree of self-organization, i.e., the extent to which community composition and functions are self-assembled or dependent on direct human inputs [@RN5445]. Novelty refers to the degree of differentiation from the historical reference communities that existed in the area before urbanization [@RN5673; @RN5674]. For example, urban communities under direct human management (e.g., park lawns) should have low naturalness and high novelty. When management stops, naturalness increases, and the community can either transition back to a state resembling the historical reference (e.g., a low novelty hay meadow) or into a novel assemblage of species (e.g., a high novelty urban grassland with non-native species) [@RN5673; @RN5674]. The fate of a specific urban site will depend on the interaction between the legacy of human intervention and inherent site characteristics [@RN2305]. Intense human legacies usually push communities into novel states, often dominated by non-native and otherwise unwanted taxa [@RN5603; @RN5604], which usually require more intensive and costly actions to restore the land for conservation purposes [@RN2305]. Understanding the patterns and drivers of urban community novelty can help to predict the restoration and rewilding potential of vacant urban spaces, facilitating their integration into the management of biodiversity in cities.

The goal of this article was to test the hypothesis that urban habitats are comprised of different degrees of ecological novelty [@RN5674]. We addressed this hypothesis by studying four urban habitats representing different land uses and human legacies, from managed park lawns to relatively unmanaged strips of land along transportation corridors, physically altered soils in residential vacant lots, and highly contaminated industrial brownfields. To address community compositional novelty, we compared the plant, fungal and bacterial species composition between urban habitats and two reference pre-urban habitats: forest remnants and hay meadows. We expected novelty to be the highest for plants, which in many cases are the object of direct management and introduction [@RN5493]; and lowest for bacteria, which should be less affected than fungi by the usually alkaline urban soils [@RN5603]. We expected novelty to be higher when urban habitats are compared to natural habitats (forests), and lower when compared to agricultural habitats (meadows). We also expected novelty to be higher in the actively managed habitat (park lawns) and in the unmanaged habitat with the highest human legacy (industrial brownfields). Finally, we expected that compositional novelty would be related to soil physicochemical properties, reflecting the major aspect of human legacy in urban habitats. To test these expectations, we tackled three main research goals: (1) to characterize the variation in soil properties across urban and pre-urban habitats; (2) to characterize the variation in plant, fungal and bacterial communities across urban and pre-urban habitats; and (3) to measure the compositional novelty across habitats and taxonomic groups.

@@

Gijón/Xixón is a medium-sized coastal city (c. 270,000 inhabitants; Instituto Nacional de Estadística) located in the Bay of Biscay, within the central Asturias metropolitan area (c. 800,000 inhabitants) of northern Spain. The city belongs to the Cantabrian Mixed Forests ecoregion [@RN5493; @RN5600], at the border between the temperate and Mediterranean climatic zones of Europe. Climate is temperate oceanic with cool summers and wet and mostly frost-free winters (mean annual temperature = 13.5 ºC, annual precipitation = 1050 mm; Agencia Estatal de Meteorología). Most of the city is built on alluvial and estuarine deposits, sand or limestone bedrock. The potential pre-human vegetation was likely dominated by deciduous mixed forests, together with riparian forests, wetlands and coastal dunes. The area has a long history of human habitation dating back to the Cantabrian Upper Paleolithic [@RN5450]. The city itself has been inhabited at least since Roman times [@RN5532]. During the medieval and early modern times, the urban center was limited to the original walled city, with fishing being the main economic activity. Expansion beyond the original core started in the 18th and 19th centuries. Exposure to global trade networks and the warm temperate climate turned the region into a hotspot for biological invasions, especially from the Americas and Asia [@RN5493]. In 1852, the city was connected by railway to regional mining and industrial centers and became a major shipping port [@RN5675]. During the following century the city experienced significant industrial development (e.g., canning, coal, steel, shipbuilding) and rapid population growth, becoming the most populated city in the Principality of Asturias [@RN5683]. Industrial and urban growth brought environmental problems including loss of natural wetlands and air and watercourse pollution [@RN5535]. Starting in the 1970s, the city and the metropolitan area have been affected by post-industrial land-use changes leaving large abandoned industrial sites (i.e., brownfields) [@RN5448; @RN5449].

@@

To survey urban biodiversity across habitats, we defined six habitat types (**Figure 1A**) representing different theoretical states of naturalness and novelty [@RN5445]: (1) forest patches \< 10 km from the city center, i.e., remnants of pre-urban natural habitats; (2) late mown hay meadows, i.e., remnants of pre-urban agricultural habitats; (3) park lawns, i.e., green spaces created and maintained by humans; (4) roadsides, i.e., relatively unmanaged strips of land alongside roads and train tracks; (5) residential vacant lots, i.e., unmanaged land which has suffered some degree of soil physical disturbance and is waiting to be urbanized; and (6) industrial vacant lots, i.e., unmanaged land with an intense legacy of soil disturbance and contamination.

## Sampling plots

During June 2023, we surveyed the city to establish five sampling plots (4 x 4 m) per each of the six habitat types (n = 30 plots). The plots were distributed across the city (**Figure 1B**) and represent the variation of vegetation types found within each habitat type. In one plot per habitat (**Figure 1B**), we buried a datalogger (M-Log5W, GeoPrecision, Ettlingen, Germany; accuracy: +/- 0.1 ºC (at 0 ºC), resolution: 0.01 ºC) at a depth of 5 cm below the upper layer of the soil. Dataloggers recorded temperature hourly for one year (1 Jan to 31 Dec 2024), after which we downloaded their records to describe the thermal environment of the city and the habitats.

```{r fig1, echo = FALSE, fig.cap = "Pre-urban and urban habitats included in this study. (A) Representative pictures of each habitat, indicating their level of naturalness (i.e., degree of ecosystem self-organization) and novelty (i.e., degree of divergence from pre-human reference ecosystems). (B) Location of the sampling plots within the city of Gijón/Xixón, Asturias, Spain, Western Europe. Each plot (dots) is colored according to habitat type. A thermometer to the right of the sampling plot indicates that a soil temperature datalogger was installed there."}
knitr::include_graphics("../results/figures/Fig1.png")
```

@@

From each of the 30 plots, we collected 500 g of soil to analyse soil physicochemical properties. We took the samples from 5 points in each plot (center + 4 cardinal points), down to a depth of 10 cm. We removed stones and other debris, sieved the soils in a 5 mm mesh and homogenized the samples in a plastic bag. Samples were analyzed at INEA Laboratory (Valladolid, Spain), following the standard methodology of the Spanish Ministry of Agriculture [@RN5599]. Soil analysis included pH, texture, bulk density, electrical conductivity, organic matter content, total N, C/N ratio, active limestone, total carbonates, assimilable P, K, Ca, Mg, Na and heavy metals (Zn, Cr, Cu, Cd, Hg, Pb, Ni).

@@

In each plot, we recorded all the spontaneous vascular plant species, as well as the cover (%) of each species in the plot. Plant species names follow Euro+Med [@RN5044]. Additionally, from each of the 30 plots, we collected soil samples for DNA metabarcoding of fungal and bacterial (including Archaea) communities. We performed the sampling in sterile conditions, and always before entering the plot for any of the other tasks. We cleaned and sterilized all tools with water and ethanol between samples. In each plot, we took samples from the four cardinal points, 0.5 m inwards from the edge, at a depth of 5 cm. Before taking the samples, we removed the vegetation with a sterile tool. Then, we collected the soil samples in a 50 ml Falcon tube, mixed them, homogenized them with a lancet and finally stored the sample in a 2 ml Eppendorf tube which we immediately froze. The frozen tubes were sent to AllGenetics & Biology SL (A Coruña, Spain) for DNA metabarcoding analysis. DNA analysis included DNA extraction; library preparation with 2 pairs of different primers specific for barcoding (ITS for fungi, and 16S for bacteria; sequencing on an Illumina NovaSeq PE250 platform (assuming an amplicon size \< 450 bp); and standard bioinformatic analysis including demultiplexing, quality control and sample preprocessing, inference of Amplicon Sequence Variants (ASVs), taxonomic assignment, number of reads per taxon and generation of rarefaction curves (see **Appendix S1** for extended DNA metabarcoding methodology).

## Statistical analysis and compositional novelty index

We performed all analyses using R version 4.3.1 [@RN5387] and the package *tidyverse* [@RN4662] for data manipulation and visualization. To describe soil temperatures as recorded by the dataloggers, we calculated the mean annual temperature, the mean diurnal range and the annual range. To assess the variability in soil physicochemical properties, we conducted a Principal Component Analysis (PCA) as implemented in the R package *FactoMineR* [@RN3166] and a correlation analysis to detect auto-correlated soil variables. Multivariate analyses of community composition were performed in *vegan* [@RN5386]: we used PERMANOVA to test differences between habitat types in their soil variables and their composition of plant species, fungal ASVs and bacterial ASVs;  Non-metric Multi-Dimensional Scaling (NMDS) ordination to assess variation in taxonomic composition of plants (abundance values of species), fungi (presence/absence of ASVs) and bacteria (presence/absence of ASVs); and environmental fitting to test the correlation between community ordination axes and soil physicochemical properties.

To calculate a compositional novelty index, we used the NMDS multidimensional distance from each urban plot (i.e., park, roadside, residential or industrial) to the centroid of the two reference pre-urban habitats (i.e., forests and meadows). We calculated compositional novelty separately for plant, fungal and bacterial communities. We tested the effect of taxonomic group, pre-urban reference habitat (forest or meadow), urban habitat (park, roadside, residential or industrial) and soil physicochemical variables on compositional novelty, by fitting a Generalized Mixed Model with the package *nlme* [@RN2333]. The model included sampling plot as a random factor. We checked and confirmed model assumptions using the package *performance* [@RN5684].

@@

## Soil properties across habitats

Soil temperature records (**Figure 2A**) indicated that forest soils were relatively buffered in their thermal range of variation, with a mean annual temperature of 13.9 ºC, a mean diurnal range of 1.8 ºC and an annual range of 10.8 ºC. Meadow and urban plots had more thermal oscillation, with a mean annual temperature between 12.4 ºC (roadsides) and 16.8 ºC (meadows), a mean diurnal range between 3.8 ºC (roadsides) and 4.6 ºC (residential lots), and an annual range between 16.5 ºC (residential lots) and 19.5 ºC (meadows).

PCA ordination of the soil physicochemical properties (**Appendix S2**) indicated that the first axis of variation, explaining 23% of the variance, ordered soils according to the concentration of heavy metals. The variables with the highest contribution to this axis (\> 10%) were the soil contents in Pb, Cu, Zn, Cd and Cr, which accounted for 66% of the variability of the first axis. This first axis did not separate urban from pre-urban habitats: for example, while some industrial lots had a very high concentration of Pb and other metals, the median Pb value of the industrial habitat was below the median value of forests (**Figure 2B**; **Appendix S2**).

The second PCA axis (**Appendix S2**) explained 21% of the variance and ordered soils according to their compaction and water retention capacities, with the highest contributing variables (\> 10%) being organic matter content, bulk density, N concentration and pH, which made up 59% of the axis' variability. This axis separated soils according to their habitat: for example, forests had high organic matter and N content, while industrial vacant lots had higher bulk density and pH (**Figure 2B**; **Appendix S2**).

The third PCA axis (**Appendix S2**) explained 10% of the variability and ordered soils according to their content in Mg, with the highest contributing variables (\>10%) being Mg content, Mg/K ratio and conductivity, together explaining 50% of the axis. In this case, like for the first axis, there was no clear separation between habitats, with industrial sites having some of the highest and lowest values (**Figure 2B**; **Appendix S2**).

The analysis showed strong correlations within the groups of variables contributing to the first, second and third PCA axes (**Appendix S2**). To avoid autocorrelation issues, for further analyses we kept three variables which taken together provide a straightforward biological interpretation of patterns in soil composition: Pb content, organic matter content and Mg content (**Figure 2B**). As previously indicated by PCA, organic matter was the only one with a clear pattern of variation: a decrease from pre-urban to urban habitats (**Figure 2B**). PERMANOVA supported that there was not a significant difference between habitats in their soil physicochemical properties (100,000 permutations, R^2^ = 0.19, F = 1.15, p = 0.31).

```{r fig2, echo = FALSE, fig.cap = "Soil properties across habitats. (A) Temperature recorded every hour in one sampling plot per habitat from 1 Jan 2024 to 31 Dec 2024, indicating the hourly records (shaded colors) and the daily averages (full colors). (B) Soil physicochemical properties. Values are shown for the three most explanatory and uncorrelated variables: soil Pb content, soil organic matter content, and soil Mg content."}
knitr::include_graphics("../results/figures/Fig2.png")
```

## Plant, fungal and bacterial communities across habitats

The survey of spontaneous plant diversity yielded 216 species, with the average richness per plot being 26 (min = 10, max = 38). Of these species, 48% were found only in urban habitats (parks, roadsides, residential lots and industrial lots), 27% occurred only in pre-urban habitats (forests and meadows) and 25% were shared by both urban and pre-urban habitats (**Fig3A**). On average, the habitat with the highest richness per plot was industrial lots (29 species), followed by residential lots (27), roadsides (24), meadows (23), parks (20) and forests (18). The most frequent species were *Holcus lanatus* (22 plots), *Plantago lanceolata* (20), *Helminthotheca echioides* (16), *Dactylis glomerata* (15) and *Lotus corniculatus* (14). PERMANOVA supported that the plant species composition was significantly different across habitats (100,000 permutations, R^2^ = 0.42, F = 3.48, p = \< 0.001). NMDS ordination of the plant species composition (**Fig. 3D**) indicated that the plots belonging to the same habitat were relatively grouped together. The major gradient in vegetation composition along the first axis was a separation between forest plots and the other habitats. Environmental fitting of the Pb, organic matter and Mg vectors onto the ordination indicated that none of the soil factors had a significant correlation (R^2^ \< 0.14, p \> 0.05) with the composition of the plant communities.
